---
title: "Data Generation : Belimumab case study (binary endpoint)" 
author: "Tristan Fauvel, Quinten Health"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cerulean
    toc: yes
    toc_float:
      collapsed: true
vignette: >
  %\VignetteIndexEntry{Data Generation : Belimumab case study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{RBExT, ggplot2}
---

```{r setup, warning=F, message=F, include=FALSE}
root <- dirname(dirname(getwd()))
knitr::opts_knit$set(root.dir = root, warning = FALSE, message = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(RBExT)
library(ggplot2)
```
 


# Belimumab case study 
```{r}
set.seed(42)
case_study_config <- yaml::yaml.load_file(system.file("conf/case_studies/belimumab.yml", package = "RBExT"))
format_case_study_config(case_study_config)
```

# Initializing a BinaryTargetData object with a normal treatment effect summary measure distribution
```{r}
source_data <- SourceData$new(case_study_config)

drift <- 0.4
target_sample_size_per_arm <- 100
```
 
 
We start by computing the rate in each arm, based on the drift in each arm (defined on the log scale).
By default, the drift in the control arm is 0.

```{r}
read_function_code(rate_from_drift_logOR)
```

```{r}
control_drift <- 0
treatment_drift <- drift + control_drift

target_data$control_rate <- rate_from_drift_logOR(control_drift, source_data$control_rate)
target_data$treatment_rate <- rate_from_drift_logOR(treatment_drift, source_data$treatment_rate)
```
## Computing the mean and standard deviation of the logOR distribution


This allows us to compute the sampling standard deviation of the log OR:

```{r}
read_function_code(standard_error_log_odds_ratio)
```

Computing the mean the logOR distribution:

```{r}
target_data$treatment_effect <- target_data$drift + source_data$treatment_effect_estimate
```

# Aggregate data generation

## Generation of aggregate data without sampling approximation

### Usage

```{r}
case_study_config$sampling_approximation <- FALSE
target_data <- TargetDataFactory$new()
target_data <- target_data$create(source_data = source_data, case_study_config = case_study_config, target_sample_size_per_arm = target_sample_size_per_arm, treatment_drift = drift, summary_measure_likelihood = source_data$summary_measure_likelihood)
```

```{r}
n_replicates <- 10000
data <- target_data$generate(n_replicates = n_replicates)
print(data[1:10,])
```
```{r}
target_data$plot_sample(data)
```

### Implementation details

```{r}
read_function_code(compute_ORs)
```

```{r}
read_function_code(sample_log_odds_ratios)
``` 
```{r}
log_OR_samples <- sample_log_odds_ratios(
  target_data$sample_size_control,
  target_data$sample_size_treatment,
  target_data$treatment_rate,
  target_data$control_rate,
  n_replicates
)

samples <- data.frame(
  treatment_effect_estimate = log_OR_samples$log_odds_ratio,
  treatment_effect_standard_error = log_OR_samples$std_err_log_odds_ratio,
  sample_size_per_arm = target_data$sample_size_per_arm
)
```

## Generation of aggregate data using a Gaussian approximation.

### Usage
```{r}
case_study_config$sampling_approximation <- TRUE

target_data <- TargetDataFactory$new()
target_data <- target_data$create(source_data = source_data, case_study_config = case_study_config, target_sample_size_per_arm = target_sample_size_per_arm, treatment_drift = drift, summary_measure_likelihood = source_data$summary_measure_likelihood)
```

```{r}
n_replicates <- 10000
data <- target_data$generate(n_replicates = n_replicates)
print(data[1:10,])
```
```{r}
target_data$plot_sample(data)
```

Notice how the sampling approximation leads to a similar distribution of the treament effect estimate, but to a very different distribution of the treatment effect standard error.


### Implementation details

```{r}
read_function_code(sample_aggregate_normal_data)
```

See [Data_generation_botox](../doc/Data_generation_botox.html) for details on aggregate normal data generation.
