---
title: "Implementation of variants of the Empirical Bayes Power Prior for normally distributed treatment effect summary measure"
author: "Tristan Fauvel, Quinten Health"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cerulean
    toc: yes
    toc_float:
      collapsed: true
vignette: >
  %\VignetteIndexEntry{Empirical Bayes Power Prior}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{RBExT, ggplot2}
---
 
## Introduction

The aim of this document is to illustrate the  implementation of variants of the Empirical Bayes Power Prior in the case where the treatment effect follows a normal distribution. 
 
## Set working directory, import functions

```{r, setwd, include=FALSE}
root <- dirname(dirname(getwd()))
knitr::opts_knit$set(root.dir = root)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

library(RBExT)
library(ggplot2)
```

## Load the Belimumab case study configuration

Load the simulation configuration and the Belimumab case study configuration from YAML files.
```{r load_config}
set.seed(42)
case_study_config <- yaml::yaml.load_file(system.file("conf/case_studies/belimumab.yml", package = "RBExT"))
```

### Create data objects

Create a source_data instance, where information about the source data is stored

 
```{r}
source_data <- ObservedSourceData$new(case_study_config)
```

Set the observed target data (in the paediatrics population)
```{r}
drift <- 0
target_data <- ObservedTargetData$new(treatment_effect_estimate = case_study_config$source$treatment_effect + drift, treatment_effect_standard_error = case_study_config$target$standard_error, target_sample_size_per_arm = as.integer(case_study_config$target$total / 2), summary_measure_likelihood = case_study_config$summary_measure_likelihood)
```

## Approach by [Gravestock et al (2017)](https://onlinelibrary.wiley.com/doi/10.1002/pst.1814)

Now, we define the model we want to use for inferring the treatment effect in the target study.  
 

```{r}
method <- "EB_PP"
# here, we choose the parameters so that the prior on the power parameter is Beta(1,1)
method_parameters <- list(
  initial_prior = "noninformative"
)

model <- Model$new()

model <- model$create(
  case_study_config = case_study_config,
  method = method,
  method_parameters = method_parameters,
  source_data = source_data
)
```

```{r}
model$plot_power_parameter_vs_drift(source_treatment_effect_estimate = source_data$treatment_effect_estimate, target_data = target_data, min_drift = -1.5, max_drift = 1.5, resolution = 100)
```

```{r}
drift <- 0
target_data <- ObservedTargetData$new(treatment_effect_estimate = case_study_config$source$treatment_effect + drift, treatment_effect_standard_error = case_study_config$target$standard_error, target_sample_size_per_arm = as.integer(case_study_config$target$total / 2), summary_measure_likelihood = case_study_config$summary_measure_likelihood)


# Perform Bayesian inference based on observed target data.
model$inference(target_data = target_data)
```

```{r}
plt <- model$plot_pdfs(xmin = 0, xmax = 1, resolution = 100)

vline_data <- data.frame(
  effect = c("Source Treatment Effect", "Target Treatment Effect"),
  xintercept = c(source_data$treatment_effect_estimate, target_data$sample$treatment_effect_estimate)
)

# Plot with vertical lines and legend
plt <- plt +
  geom_vline(aes(xintercept = xintercept, color = effect, linetype = effect),
    data = vline_data, size = 1
  )

# Customize the legend (optional)
plt + scale_color_manual(name = "", values = c("Source Treatment Effect" = "blue", "Target Treatment Effect" = "red")) +
  scale_linetype_manual(name = "", values = c("Source Treatment Effect" = "dashed", "Target Treatment Effect" = "dotted")) +
  theme_minimal()

print(plt)
```
 
With more drift : 

```{r}
drift <- -2
target_data <- ObservedTargetData$new(treatment_effect_estimate = case_study_config$source$treatment_effect + drift, treatment_effect_standard_error = case_study_config$target$standard_error, target_sample_size_per_arm = as.integer(case_study_config$target$total / 2), summary_measure_likelihood = case_study_config$summary_measure_likelihood)
model$inference(target_data = target_data)

plt <- model$plot_pdfs(xmin = 0 + drift, xmax = case_study_config$source$treatment_effect, resolution = 100)

vline_data <- data.frame(
  effect = c("Source Treatment Effect", "Target Treatment Effect"),
  xintercept = c(source_data$treatment_effect_estimate, target_data$sample$treatment_effect_estimate)
)

# Plot with vertical lines and legend
plt <- plt +
  geom_vline(aes(xintercept = xintercept, color = effect, linetype = effect),
    data = vline_data, size = 1
  )

# Customize the legend (optional)
plt + scale_color_manual(name = "", values = c("Source Treatment Effect" = "blue", "Target Treatment Effect" = "red")) +
  scale_linetype_manual(name = "", values = c("Source Treatment Effect" = "dashed", "Target Treatment Effect" = "dotted"))

print(plt)
```
This illustrates the lack of adaptiveness of the Empirical Bayes Power Prior as proposed by Gravestock et al, as a large discrepancy between the prior and the likelihood still results in a strong weight of source data.

 

## p-value based Power Prior

In a generalization of the test-then-pool approach, [Liu et al (2018)](https://onlinelibrary.wiley.com/doi/10.1002/pst.1836) proposed a method for selecting the power parameter $\gamma$ in the conditional power prior based on the p-value of an equivalence test between the source and target data. The equation used to determine $\gamma$ is as follows:
\(\gamma = \exp\left[\frac{k}{1-p}\ln(1-p) \right]\),

where $k$ is a shape parameter that must be specified. This function was chosen so that more source data is borrowed when the p-value is close to 0 (i.e., the non-equivalence null hypothesis is strongly rejected). Larger values of $k$ result in steeper curves (faster decrease from 0 to 1), that is more discounting will be applied to the source data for a given p-value. This method can be viewed as an extension of the test-then-pool approach, with the power parameter smoothly adjusting the amount of borrowing from no borrowing to pooling. 

```{r}
method <- "p_value_based_PP"
# here, we choose the parameters so that the prior on the power parameter is Beta(1,1)
method_parameters <- list(
  initial_prior = "noninformative",
  shape_parameter = 0.5,
  equivalence_margin = 0.1
)

model <- Model$new()

model <- model$create(
  case_study_config = case_study_config,
  method = method,
  method_parameters = method_parameters,
  source_data = source_data
)
```
  


```{r}
model$plot_power_parameter_vs_drift(source_treatment_effect_estimate = source_data$treatment_effect_estimate, target_data = target_data, min_drift = -0.15, max_drift = 0.15, resolution = 100)
```

```{r}
drift <- 0
target_data <- ObservedTargetData$new(treatment_effect_estimate = case_study_config$source$treatment_effect + drift, treatment_effect_standard_error = case_study_config$target$standard_error, target_sample_size_per_arm = as.integer(case_study_config$target$total / 2), summary_measure_likelihood = case_study_config$summary_measure_likelihood)

# Perform Bayesian inference based on observed target data.
model$inference(target_data = target_data)
```
Posterior value of the power prior parameter:

```{r}
model$posterior_parameters
```

```{r}
model$plot_pdfs(xmin = -1, xmax = 1, resolution = 100, target_data)
```

With more drift : 

```{r}
drift <- -2
target_data <- ObservedTargetData$new(treatment_effect_estimate = case_study_config$source$treatment_effect + drift, treatment_effect_standard_error = case_study_config$target$standard_error, target_sample_size_per_arm = as.integer(case_study_config$target$total / 2), summary_measure_likelihood = case_study_config$summary_measure_likelihood)
model$inference(target_data = target_data)

plt <- model$plot_pdfs(xmin = 0 + drift, xmax = case_study_config$source$treatment_effect, resolution = 100)

vline_data <- data.frame(
  effect = c("Source Treatment Effect", "Target Treatment Effect"),
  xintercept = c(source_data$treatment_effect_estimate, target_data$sample$treatment_effect_estimate)
)

# Plot with vertical lines and legend
plt <- plt +
  geom_vline(aes(xintercept = xintercept, color = effect, linetype = effect),
    data = vline_data, size = 1
  )

# Customize the legend (optional)
plt + scale_color_manual(name = "", values = c("Source Treatment Effect" = "blue", "Target Treatment Effect" = "red")) +
  scale_linetype_manual(name = "", values = c("Source Treatment Effect" = "dashed", "Target Treatment Effect" = "dotted"))

print(plt)
```

## PDCCPP

Now, we define the model we want to use for inferring the treatment effect in the target study.  
 

```{r}
method <- "PDCCPP"
# here, we choose the parameters so that the prior on the power parameter is Beta(1,1)
method_parameters <- list(
  initial_prior = "noninformative",
  desired_tie = 0.065,
  significance_level = 0.05,
  tolerance = 0.0001,
  n_iter = 1e6
)

model <- Model$new()

model <- model$create(
  case_study_config = case_study_config,
  method = method,
  method_parameters = method_parameters,
  source_data = source_data
)
```

```{r}
model$plot_power_parameter_vs_drift(source_treatment_effect_estimate = source_data$treatment_effect_estimate, target_data = target_data, min_drift = -0.5, max_drift = 0.5, resolution = 100)
```

```{r}
drift <- 0
target_data <- ObservedTargetData$new(treatment_effect_estimate = case_study_config$source$treatment_effect + drift, treatment_effect_standard_error = case_study_config$target$standard_error, target_sample_size_per_arm = as.integer(case_study_config$target$total / 2), summary_measure_likelihood = case_study_config$summary_measure_likelihood)


# Perform Bayesian inference based on observed target data.
model$inference(target_data = target_data)
```
```{r}
model$plot_pdfs(xmin = 0, xmax = 1, resolution = 100)
```

With more drift : 

```{r}
drift <- -2
target_data <- ObservedTargetData$new(treatment_effect_estimate = case_study_config$source$treatment_effect + drift, treatment_effect_standard_error = case_study_config$target$standard_error, target_sample_size_per_arm = as.integer(case_study_config$target$total / 2), summary_measure_likelihood = case_study_config$summary_measure_likelihood)
model$inference(target_data = target_data)

plt <- model$plot_pdfs(xmin = 0 + drift, xmax = case_study_config$source$treatment_effect, resolution = 100)

vline_data <- data.frame(
  effect = c("Source Treatment Effect", "Target Treatment Effect"),
  xintercept = c(source_data$treatment_effect_estimate, target_data$sample$treatment_effect_estimate)
)

# Plot with vertical lines and legend
plt <- plt +
  geom_vline(aes(xintercept = xintercept, color = effect, linetype = effect),
    data = vline_data, size = 1
  )

# Customize the legend (optional)
plt + scale_color_manual(name = "", values = c("Source Treatment Effect" = "blue", "Target Treatment Effect" = "red")) +
  scale_linetype_manual(name = "", values = c("Source Treatment Effect" = "dashed", "Target Treatment Effect" = "dotted"))

print(plt)
```
