---
title: "Normalized Power Prior : normal likelihood, beta prior on the power parameter, noninformative initial prior, known sampling variance"
author: "Tristan Fauvel, Quinten Health"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cerulean
    toc: yes
    toc_float:
      collapsed: true
vignette: >
  %\VignetteIndexEntry{Introduction to MyPackage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{RBExT, ggplot2}
---

## Introduction

The aim of this document is to illustrate the  implementation of the Normalized Power Prior in the case where the treatment effect follows a normal distribution. 

The derivation for the normalized power prior can be found in [Pawel et al (2023)](https://doi.org/10.1002/sta4.591)
 
With known standard error $\sigma_T$ of the estimate $\theta_T$, a Beta prior on the power parameter $\gamma \sim \operatorname{Be}(p, q)$. This choice leads to the normalized power prior : 

$$
\pi\left(\theta, \gamma \mid D_S\right)=\frac{\mathcal{L}\left(D_S \mid \theta\right)^\gamma \pi(\gamma)}{\int_{-\infty}^{+\infty} \mathcal{L}\left(D_S \mid \theta^{\prime}\right)^\gamma d \theta^{\prime}}=\mathcal{N}\left(\theta \mid \hat{\theta}_S, \sigma_S^2 / \gamma\right) \operatorname{Be}(\gamma \mid p, q)
$$

Combining this prior with the likelihood of the target study data produces a joint posterior for $\theta$ and $\gamma$, that is,
$$
\pi\left(\theta, \gamma \mid D_T, D_S\right)=\frac{L(D_T \mid \theta) \pi\left(\theta, \gamma \mid D_S\right)}{\int_0^1 \int_{-\infty}^{\infty} L\left(D_T \mid \theta^{\prime}\right) \pi\left(\theta^{\prime}, \gamma^{\prime} \mid D_S\right) d \theta^{\prime} d \gamma^{\prime}}=\frac{\mathcal{N}\left(\hat{\theta}_T \mid \theta, \sigma_T^2\right) \mathcal{N}\left(\theta \mid \hat{\theta}_S, \sigma_S^2 / \gamma\right) \operatorname{Be}(\gamma \mid p, q)}{\int_0^1 \mathcal{N}\left(\hat{\theta}_T \mid \hat{\theta}_S, \sigma_T^2+\sigma_S^2 / \gamma^{\prime}\right) \operatorname{Be}\left(\gamma^{\prime} \mid p, q\right) d \gamma^{\prime}},
$$
from which a marginal posterior for $\gamma$ can be obtained by integrating out $\theta$, that is,
$$
\pi\left(\gamma \mid D_T, D_S\right)=\int_{-\infty}^{+\infty} \pi\left(\theta, \gamma \mid D_T, D_S\right) d \theta=\frac{\mathcal{N}\left(\hat{\theta}_T \mid \hat{\theta}_S, \sigma_T^2+\sigma_S^2 / \gamma\right) \operatorname{Be}(\gamma \mid p, q)}{\int_0^1 \mathcal{N}\left(\hat{\theta}_T \mid \hat{\theta}_S, \sigma_T^2+\sigma_S^2 / \gamma^{\prime}\right) \operatorname{Be}\left(\gamma^{\prime} \mid p, q\right) d \gamma^{\prime}} .
$$

## Set working directory, import functions

```{r, setwd, include=FALSE}
root <- dirname(dirname(getwd()))
knitr::opts_knit$set(root.dir = root)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

library(RBExT)
library(ggplot2)
```

## Load the Belimumab case study configuration

Load the simulation configuration and the Belimumab case study configuration from YAML files.
```{r load_config}
set.seed(42)
case_study_config <- yaml::yaml.load_file(system.file("conf/case_studies/belimumab.yml", package = "RBExT"))
```

### Create data objects

Create a source_data instance, where information about the source data is stored

 
```{r}
source_data <- ObservedSourceData$new(case_study_config)
```

Set the observed target data (in the paediatrics population)
```{r}
target_data <- ObservedTargetData$new(treatment_effect_estimate = case_study_config$target$treatment_effect, treatment_effect_standard_error = case_study_config$target$standard_error, target_sample_size_per_arm = as.integer(case_study_config$target$total / 2), summary_measure_likelihood = case_study_config$summary_measure_likelihood)
```

## Create the model

Now, we define the model we want to use for inferring the treatment effect in the target study.  

We reuse the code published in [Pawel et al (2023)](https://doi.org/10.1002/sta4.591). Marginal posterior densities are computed by numerical integration:

```{r}
method <- "NPP"
# here, we choose the parameters so that the prior on the power parameter is Beta(1,1)
method_parameters <- list(
  initial_prior = "noninformative",
  power_parameter_mean = 0.5,
  power_parameter_std = sqrt(1 / 12)
)

model <- Model$new()

model <- model$create(
  case_study_config = case_study_config,
  method = method,
  method_parameters = method_parameters,
  source_data = source_data
)
```
   
### Inference

 

```{r}
# Perform Bayesian inference based on observed target data.
model$inference(target_data = target_data)
```


```{r}
model$plot_pdfs(xmin = 0, xmax = 1, resolution = 100)
```

Posterior distribution of the power parameter:
```{r}
model$plot_power_parameter_posterior_pdf(target_data)
```
## Power parameter as a function of drift
```{r}
target_treatment_effect_values <- seq(-3, 3, length.out = 100)

drift <- target_treatment_effect_values - source_data$treatment_effect_estimate

posterior_power_parameter_means <- numeric(100)
posterior_power_parameter_std <- numeric(100)

for (i in 1:length(target_treatment_effect_values)) {
  target_data$sample$treatment_effect_estimate <- target_treatment_effect_values[i]
  model$inference(target_data)
  posterior_power_parameter_means[i] <- model$posterior_parameters$power_parameter_mean
  posterior_power_parameter_std[i] <- model$posterior_parameters$power_parameter_std
}
```

```{r}
# Combine into a data frame
plot_data <- data.frame(
  drift = drift,
  target_treatment_effect = target_treatment_effect_values,
  power_parameter_mean = posterior_power_parameter_means,
  power_parameter_std = posterior_power_parameter_std
)
 
ggplot(plot_data, aes(x = drift, y = power_parameter_mean)) +
  geom_point() +
  # geom_vline(xintercept = source_data$treatment_effect_estimate, linetype = "dashed", color = "red", linewidth = 1, show.legend = TRUE) +
  # geom_point(aes(x = source_data$treatment_effect_estimate, y = Inf, color = "Source Treatment Effect"), shape = "|", size = 5) +
  # scale_color_manual(name = "", values = c("Source Treatment Effect" = "red")) +
  labs(
    title = "Posterior Mean of the Power Parameter PDF vs Drift",
    x = "Drift",
    y = "Power Parameter Mean"
  ) +
  theme_minimal()
```

```{r}
ggplot(plot_data, aes(x = drift, y = power_parameter_std)) +
  geom_point() +
  labs(
    title = "Posterior STD of the Power Parameter PDF vs Drift",
    x = "Drift",
    y = "Power Parameter STD"
  ) +
  theme_minimal()
```
