---
title: "Implementation of the PDCCPP"
author: "Tristan Fauvel, Quinten Health"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cerulean
    toc: yes
    toc_float:
      collapsed: true
vignette: >
  %\VignetteIndexEntry{Introduction to MyPackage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{RBExT, ggplot2}
---

## Introduction

The aim of this document is to illustrate the PDCCPP ([Nikolakopoulos et al (2018)](https://onlinelibrary.wiley.com/doi/10.1111/biom.12835)), the empirical Bayes power prior (as proposed by [Gravestock et al (2017)](https://onlinelibrary.wiley.com/doi/10.1002/pst.1814)), and the p-value based power prior ([Liu et al, (2018)](https://pubmed.ncbi.nlm.nih.gov/29125220/)).


## Set working directory, import functions

```{r setup, warning=F, message=F, include = FALSE}
root <- dirname(dirname(getwd()))
knitr::opts_knit$set(root.dir = root, warning = FALSE, message = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

library(RBExT)
library(ggplot2)
```

## Load the Belimumab case study configuration

Load the simulation configuration and the Belimumab case study configuration from YAML files.

```{r load_config}
set.seed(42)
config_path <- system.file("conf/full/simulation_config.yml", package = "RBExT")
simulation_config <- yaml::yaml.load_file(config_path)

config_path <- system.file("conf/case_studies/aprepitant.yml", package = "RBExT")
case_study_config <- yaml::yaml.load_file(config_path)
```
 
  

## Create data objects

Create a source_data instance, where information about the source data is stored

```{r}
target_sample_size_per_arm <- as.integer(case_study_config$target$total / 2)
drift <- 0.4

source_data <- ObservedSourceData$new(case_study_config)
print(source_data)
```

Set the observed target data (in the paediatrics population)
```{r}
summary_measure_likelihood <- case_study_config$summary_measure_likelihood
endpoint <- case_study_config$endpoint
target_data <- ObservedTargetData$new(treatment_effect_estimate = case_study_config$target$treatment_effect, treatment_effect_standard_error = case_study_config$target$standard_error, target_sample_size_per_arm = as.integer(case_study_config$target$total / 2), summary_measure_likelihood = case_study_config$summary_measure_likelihood)
```

## PDCCPP 


We note that the PDCCPP method was developed in the Gaussian case only, assuming that the sampling standard deviation is known and is the same in the source and target studies. However, in our simulation study, we do not make this latter assumption. To reuse the code by [Nikolakopoulos et al (2018)](https://onlinelibrary.wiley.com/doi/10.1111/biom.12835),  what we adapt the sample size per arm in the source study by replacing it with an "effective sample size per arm at variance $\sigma_T^2$, $N_0$: 
We set: 
$N_0 = N_S\frac{\sigma_T^2}{\sigma_S^2}$
So that everything is equivalent to the case where the sampling std is $\sigma_T$ in the target study and the source study, but with a sample size per arm $N_0$ in the source study (instead of $N_S$)


```{r}
method <- "PDCCPP"
method_parameters <- list(
  initial_prior = "noninformative", # This corresponds to the fact that the posterior for the adults data is derived from an uninformative prior (for consistency with other methods). May be removed in future releases.
  empirical_bayes = FALSE # Corresponds to whether some parameters of the prior are based on observed data.
)
```

Now, we define the model we want to use for inferring the treatment effect in the target study.  

```{r}
model <- Model$new()

model <- model$create(
  case_study_config = case_study_config,
  method = method,
  method_parameters = method_parameters,
  source_data = source_data
)
``` 


### Inference with the PDCCPP

To perform inference, we would simply use: 
```{r}
# Perform Bayesian inference based on observed target data.
model$inference(target_data = target_data)
```  

This inference steps set the following attributes which, taken together, specify the posterior distribution.
```{r}
print(model$wpost) # Posterior weight
print(model$vague_posterior_mean)
print(model$vague_posterior_variance)
print(model$info_posterior_mean)
print(model$info_posterior_variance)
```

Now, let us see how this is implemented:

```{r}
read_function_code(model$inference)
```

The first step in inference is to update the prior based on target_data if needed, in case where the method uses empirical Bayes. With empirical Bayes, some parameters of the prior are set based on observed target data. 
The second step is to compute the moments of the treatment effect posterior distribution. Note that we use the same inference function for all methods/scenarios combinations that do not require MCMC (in which case we need samples from the posterior and not only the parameters of the posterior). 


In the study protocol, we specified that the variance of the vague component of the RMP should be such that this prior corresponds to the information provided by a single subject per arm in the target study. This is a form of empirical Bayes:   

```{r}
read_function_code(model$empirical_bayes_update)
```
 

The posterior is given by : 
 
$$p(\theta_T | \boldsymbol{y}_S, \boldsymbol{y}_T) = \widetilde{w}\mathcal{N}\left(\theta_T \bigg| \frac{\mu_v}{N_T\frac{\sigma^2_v}{\sigma^2_T} + 1} + \frac{\overline{y}_T}{\frac{\sigma^2_T}{N_T\sigma_v^2} + 1}, \left(\sigma_v^{-2} + \frac{N_T}{\sigma^2_T} \right)^{-1}\right) + (1- \widetilde{w})\mathcal{N}\left(\theta_T \bigg| \frac{\overline{y}_S}{N_T\frac{\nu^2_S}{\sigma^2_T} + 1} + \frac{\overline{y}_T}{\frac{\sigma^2_T}{N_T\nu_S^2} + 1}, \left(\nu_S^{-2} + \frac{N_T}{\sigma^2_T} \right)^{-1}\right),
$$

where $\mathcal{N}(\theta_T | \mu, \sigma^2)$ denotes the probability density function of a normal distribution with mean $\mu$ and variance $\sigma^2$, evaluated at $\theta_T$.


The code to compute the posterior moments is the following : 

```{r}
read_function_code(model$posterior_moments)
```
 
With :

```{r}
read_function_code(model$posterior_mean)
```

and :  
```{r}
read_function_code(model$posterior_variance)
```

The variance of a mixture of two distributions is given by : 

```{r}
read_function_code(model$mixture_variance)
```

A crucial aspect of this inference step is the computation of the posterior mixture weights:

The posterior weight of the vague component $\widetilde{w}$ is given by:

$$
    \tilde{w} = \frac{wC_v}{wC_v + (1-w)C_S}
$$
Where $C_v$ and $C_S$ are proportional to the marginal likelihood (or prior predictive probability) of the aggregate data for each Gaussian component:
$$C_v =  \frac{1}{\sqrt{\sigma_v^2 + \sigma^2_T/N_T}}\exp\left(-\frac{1}{2}\frac{(\overline{y}_T - \mu_v)^2}{\sqrt{\sigma_v^2 + \sigma^2_T/N_T}}\right)$$
and : 
$$C_S =  \frac{1}{\sqrt{\nu_S^2 + \sigma^2_T/N_T}}\exp\left(-\frac{1}{2}\frac{(\overline{y}_T - \overline{y}_S)^2}{\sqrt{\nu_S^2 + \sigma^2_T/N_T}}\right)$$
 

```{r}
read_function_code(posterior_weight)
```
```{r}
read_function_code(posterior_pdf)
```

```{r}
read_function_code(posterior_cdf)
```


Let us plot the posterior pdf : 
 

```{r}
library(ggplot2)
x_values <- seq(-0.5, 2, by = 0.01) # Range of values for x

prior_pdf <- model$prior_pdf(x_values)
posterior_pdf <- model$posterior_pdf(x_values)

df <- data.frame(x = x_values, prior_pdf = prior_pdf, posterior_pdf = posterior_pdf)

# Use ggplot2 to plot both PDFs on the same plot
ggplot2::ggplot(df, ggplot2::aes(x = x)) +
  geom_line(ggplot2::aes(y = prior_pdf), color = "blue", size = 1) +
  geom_line(ggplot2::aes(y = posterior_pdf), color = "red", size = 1) +
  ggplot2::labs(
    title = "Prior and posterior distribution of the treatment effect",
    x = "Value",
    y = "Density",
    labels = c("Prior pdf", "Posterior pdf")
  ) +
  theme_minimal() +
  scale_color_manual(values = c("blue", "red")) +
  guides(color = guide_legend(title = NULL)) +
  scale_fill_manual(
    name = "PDF of the robust mixture of Gaussians",
    labels = c("Prior pdf", "Posterior pdf"),
    values = c("blue", "red")
  )
```                    

Being able to sample from the posterior is crucial for estimating quantiles, we use the following:. 

```{r}
read_function_code(sample_posterior)
```

Illustration of this sampling approach:

```{r}
samples <- model$sample_posterior(10000)
hist(samples, breaks = 60, col = "skyblue", main = "Samples from the posterior distribution of the treatment effect", xlab = "Treatment effect", ylab = "Number of samples")
```

```{r}
sessionInfo()
```
 
