---
title: "Pooling"
author: "Tristan Fauvel, Quinten Health"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cerulean
    toc: yes
    toc_float:
      collapsed: yes  
  pdf_document:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Pooling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{RBExT, ggplot2}
---

  
## Introduction

This vignette describes an implementation of a Bayesian analysis that pools the source and target study data.

## Posterior distribution 

With pooling, the prior distribution of $\theta_T$ is given by:

\[ p(\theta_T) = \mathcal{N}(\theta_T | \hat{\theta}_S, \sigma^2_S/N_S) \]

We observe \(N_T\) data points with sample mean \(\hat{\theta}_T\) and known variance \(\sigma^2_T\). This implies the likelihood is given by:

\[ p(\hat{\theta}_T | \theta_T) = \mathcal{N}(\hat{\theta}_T | \theta_T, \sigma^2_T/N_T) \]


Since both the prior and the likelihood are Gaussian distributions, the posterior will also be a Gaussian distribution.

The posterior distribution \( p(\theta_T | \hat{\theta}_T) \) is :

\[ \theta_T | \hat{\theta}_T \sim \mathcal{N}(\mu_{\text{post}}, \sigma^2_{\text{post}}) \]

**Posterior Mean:**

\[ \mu_{\text{post}} = \sigma_{\text{post}}^2 \left( \frac{\hat{\theta}_S N_S}{\sigma^2_S} + \frac{\hat{\theta}_T N_T}{\sigma^2_T} \right) \]

 **Posterior Variance:**

\[ \sigma_{\text{post}}^2 = \left( \frac{N_S}{\sigma^2_S} + \frac{N_T}{\sigma^2_T} \right)^{-1} \]


Thus, the posterior distribution \( p(\theta_T | \hat{\theta}_T) \) is:

\[ \theta_T | \hat{\theta}_T \sim N\left( \left( \frac{N_S}{\sigma^2_S} + \frac{N_T}{\sigma^2_T} \right)^{-1} \left( \frac{\hat{\theta}_S N_S}{\sigma^2_S} + \frac{\hat{\theta}_T N_T}{\sigma^2_T} \right), \left( \frac{N_S}{\sigma^2_S} + \frac{N_T}{\sigma^2_T} \right)^{-1} \right) \]


In the limit where \( \sigma_T \) goes to infinity:

\[ \sigma_{\text{post}}^2 \rightarrow \frac{\sigma^2_S}{N_S} \]

\[ \mu_{\text{post}} \rightarrow \hat{\theta}_S \]

That is, the target study data are discarded because the likelihood becomes flat, and the posterior is the same as the prior.



Pooling the data is equivalent to defining a Gaussian prior with mean and standard deviation corresponding to the treatment effect estimate and standard error on the treatment effect in the source study.
 
  
## Set working directory and load packages

```{r setup, warning=F, message=F, include = FALSE}
root <- dirname(dirname(getwd()))
knitr::opts_knit$set(root.dir = root, warning = FALSE, message = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

library(RBExT)
library(ggplot2)
```

## Load the case study configuration

Load the simulation configuration and the Belimumab case study configuration from YAML files.

```{r load_config}
set.seed(42)
case_study_config <- yaml::yaml.load_file(system.file("conf/case_studies/belimumab.yml", package = "RBExT"))
```

### Create data objects

Create a source_data instance, where information about the source data is stored

 
```{r}
source_data <- ObservedSourceData$new(case_study_config)
```

Set the observed target data (in the paediatrics population)
```{r}
target_data <- ObservedTargetData$new(treatment_effect_estimate = case_study_config$target$treatment_effect, treatment_effect_standard_error = case_study_config$target$standard_error, target_sample_size_per_arm = as.integer(case_study_config$target$total / 2), summary_measure_likelihood = case_study_config$summary_measure_likelihood)
```

### Create model
  
 
```{r}
method <- "pooling"
method_parameters <- list(
  initial_prior = "noninformative" # This corresponds to the fact that the posterior for the adults data is derived from an uninformative prior (for consistency with other methods).
)
``` 
Now, we define the model we want to use for inferring the treatment effect in the target study.  

```{r}
model <- Model$new()

model <- model$create(
  case_study_config = case_study_config,
  method = method,
  method_parameters = method_parameters,
  source_data = source_data
)
```

This model calls RBesT in the backend, leveraging the fact that it is a special case of a Robust Mixture Prior.
   
### Inference


```{r}
# Perform Bayesian inference based on observed target data.
model$inference(target_data = target_data)

model$plot_pdfs(xmin = 0, xmax = 1, resolution = 100)
```
