---
title: "Separate analysis"
author: "Tristan Fauvel, Quinten Health"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cerulean
    toc: yes
    toc_float:
      collapsed: yes  
  pdf_document:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Separate analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{RBExT, ggplot2}
---


## Introduction

This vignette describes an implementation of a Bayesian analysis without borrowing.

  
## Set working directory and load packages

```{r setup, warning=F, message=F, include = FALSE}
root <- dirname(dirname(getwd()))
knitr::opts_knit$set(root.dir = root, warning = FALSE, message = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

library(RBExT)
library(ggplot2)
```

## Load the case study configuration

Load the simulation configuration and the Belimumab case study configuration from YAML files.

```{r load_config}
set.seed(42)
case_study_config <- yaml::yaml.load_file(system.file("conf/case_studies/belimumab.yml", package = "RBExT"))
```

### Create data objects

Create a source_data instance, where information about the source data is stored

 
```{r}
source_data <- ObservedSourceData$new(case_study_config)
```

Set the observed target data (in the paediatrics population)
```{r}
target_data <- ObservedTargetData$new(treatment_effect_estimate = case_study_config$target$treatment_effect, treatment_effect_standard_error = case_study_config$target$standard_error, target_sample_size_per_arm = as.integer(case_study_config$target$total / 2), summary_measure_likelihood = case_study_config$summary_measure_likelihood)
```

### Create model
  
 
```{r}
method <- "separate"
method_parameters <- list(
  initial_prior = "noninformative" # This corresponds to the fact that the posterior for the adults data is derived from an uninformative prior (for consistency with other methods).
)
``` 
Now, we define the model we want to use for inferring the treatment effect in the target study.  

```{r}
model <- Model$new()

model <- model$create(
  case_study_config = case_study_config,
  method = method,
  method_parameters = method_parameters,
  source_data = source_data
)
```
   
### Inference


```{r}
# Perform Bayesian inference based on observed target data.
model$inference(target_data = target_data)

model$plot_pdfs(xmin = -5, xmax = 5, resolution = 100)
```
