---
title: "Replication of the Belimumab case study"
author: "Tristan Fauvel, Quinten Health"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cerulean
    toc: yes
    toc_float:
      collapsed: true
vignette: >
  %\VignetteIndexEntry{Replication of the Belimumab case study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{RBExT, ggplot2}
---

## Introduction

This code replicates the analysis presented in [Best et al (2023)](https://www.tandfonline.com/doi/full/10.1080/19466315.2024.2342817), initially reported in [Pottackal et al (2019)](https://acrabstracts.org/abstract/application-of-bayesian-statistics-to-support-approval-of-intravenous-belimumab-in-children-with-systemic-lupus-erythematosus-in-the-united-states/).

## Set working directory, import functions and configurations

```{r, setwd, include=FALSE}
root <- dirname(dirname(getwd()))
knitr::opts_knit$set(root.dir = root)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

library(RBExT)
library(ggplot2)
```


Load the Belimumab case study configuration from YAML files.

```{r load_config}
set.seed(42)

case_study_config <- yaml::yaml.load_file(system.file("conf/case_studies/belimumab.yml", package = "RBExT"))
```


## Define the method used (RMP) and the range of parameters considered


Define the range of values considered for the weight on the informative component of the RMP
```{r parameters_range}
method <- "RMP"
Nw <- 21
w_range <- seq(0, 1, length.out = Nw)
```

Define the method's parameters
```{r parameters}
method_parameters <- list(
  initial_prior = "noninformative", # This corresponds to the fact that the posterior for the adults data is derived from an uninformative prior (for consistency with other methods). May be removed in future releases.
  prior_weight = 0.5, # weight on the informative component of the mixture
  empirical_bayes = TRUE
)
```


## Create data objects

Create a source_data instance, where information about the source data is stored

```{r}
source_data <- ObservedSourceData$new(case_study_config)
```

Set the observed target data (in the paediatric population)
```{r}
target_data <- ObservedTargetData$new(treatment_effect_estimate = case_study_config$target$treatment_effect, treatment_effect_standard_error = case_study_config$target$standard_error, target_sample_size_per_arm = as.integer(case_study_config$target$total / 2), summary_measure_likelihood = case_study_config$summary_measure_likelihood)
```
Note that, at the moment, the code does not allow specifying different sample sizes for the arms of the target study.
 

## Inference

For each parameter, instantiate a model, perform Bayesian inference and compute the posterior mean and credible interval.

```{r init}
point_est <- numeric(Nw)
credible_intervals <- matrix(NA, 2, Nw)
confidence_level <- 0.95
lower_quantile <- (1 - confidence_level) / 2
upper_quantile <- 1 - lower_quantile

posterior_w <- numeric(Nw)
```

The model instantiation relies on the so-called factory design pattern for increased ease of use and flexibility.

```{r frequentist_ocs, message=FALSE, error=FALSE}
for (i in seq_along(w_range)) {
  method_parameters$prior_weight <- w_range[i]

  model <- Model$new()
  model <- model$create(
    case_study_config = case_study_config,
    method = method,
    method_parameters = method_parameters,
    source_data = source_data
  )

  # Perform Bayesian inference based on observed target data.
  model$inference(target_data = target_data)

  # Convert the estimates back to the natural scale
  point_est[i] <- exp(model$posterior_mean())

  # To estimate the credible intervals, we sample from the posterior distribution
  posterior_samples <- exp(model$sample_posterior(10000))
  credible_intervals[, i] <- quantile(posterior_samples, probs = c(lower_quantile, upper_quantile))
  posterior_w[i] <- model$wpost
}
```

## Plot the results

We plot the results as in the original paper by [Pottackal et al (2019)](https://acrabstracts.org/abstract/application-of-bayesian-statistics-to-support-approval-of-intravenous-belimumab-in-children-with-systemic-lupus-erythematosus-in-the-united-states/).

```{r plot_results, fig.width=5, fig.height=3}
ggplot2::ggplot() +
  geom_errorbar(ggplot2::aes(x = w_range, y = point_est, ymin = credible_intervals[1, ], ymax = credible_intervals[2, ]),
    color = "black", linewidth = 1, width = 0
  ) +
  geom_point(ggplot2::aes(x = w_range, y = point_est), color = "black") +
  geom_hline(yintercept = 1, linetype = "dashed", linewidth = 1) +
  ggplot2::labs(
    x = "Prior weight on adult data",
    y = "Odds ratio (95% CrI, log scale)",
    title = "Posterior Mean with the Gaussian RMP"
  ) +
  scale_y_log10(breaks = c(0.7, 0.9, 1, 1.5, 2, 3)) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

Posterior and prior pdf : 
```{r}
method_parameters$prior_weight <- 0.5

model <- Model$new()
model <- model$create(
  case_study_config = case_study_config,
  method = method,
  method_parameters = method_parameters,
  source_data = source_data
)

# Perform Bayesian inference based on observed target data.
model$inference(target_data = target_data)
model$plot_pdfs(xmin = -1, xmax = 1, resolution = 100)
```
