---
title: "Replication of the Mepolizumab case study"
author: "Tristan Fauvel, Quinten Health"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cerulean
    toc: yes
    toc_float:
      collapsed: true
resource_files:
  - images/RMP_binary.png
vignette: >
  %\VignetteIndexEntry{Replication of the Mepolizumab case study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{RBExT, ggplot2}
---


## Introduction

This code replicates the analysis presented in [Best et al (2021)](https://onlinelibrary.wiley.com/doi/10.1002/pst.2093).

### Set working directory, import functions and configurations

```{r, setwd, include=FALSE}
root <- dirname(dirname(getwd()))
knitr::opts_knit$set(root.dir = root)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

library(RBExT)
library(ggplot2)
library(RBesT)
library(dplyr)
library(kableExtra)
```
 
 
Load the Mepolizumab case study configuration from YAML files.

```{r load_config}
set.seed(42)
config_path <- system.file("conf/case_studies/mepolizumab.yml", package = "RBExT")
case_study_config <- yaml::yaml.load_file(config_path)
```


### Define the method used (RMP) and the range of parameters considered


Define the range of values considered for the weight on the informative component of the RMP
```{r parameters_range}
method <- "RMP"
Nw <- 20
w_range <- seq(0, 1, length.out = Nw)
```

Define the method's parameters
```{r parameters}
method_parameters <- list(
  initial_prior = "noninformative", # This corresponds to the fact that the posterior for the adults data is derived from an uninformative prior (for consistency with other methods). May be removed in future releases.
  prior_weight = 0.5, # weight on the informative component of the mixture
  empirical_bayes = FALSE
)
```


Set the vague component variance using the target data variance (as in  [Best et al (2021)](https://onlinelibrary.wiley.com/doi/10.1002/pst.2093) for Mepolizumab). The vague prior corresponds to the information provided by a single subject.

```{r target_data}
vague_prior_variance <- case_study_config$target$standard_error^2 * case_study_config$target$total
# This is obtained by setting
empirical_bayes <- TRUE
```

<!-- 2nd option: -->

<!-- Set the vague component variance using the source data variance (as in Best et al, 2023, for Belimumab). The vague prior corresponds to the information provided by a single subject per arm in the source study -->

<!-- ```{r source_data} -->
<!-- vague_prior_variance <- case_study_config$source$standard_error^2 * case_study_config$source$total / 2 -->
<!-- # This is obtained by setting -->
<!-- empirical_bayes <- FALSE -->

<!-- method_parameters <- list( -->
<!--   initial_prior = "noninformative", -->
<!--   prior_weight = 0.5, -->
<!--   empirical_bayes = empirical_bayes -->
<!-- ) -->
<!-- ``` -->


### Create data objects

Create a source_data instance, where information about the source data is stored

 
```{r}
source_data <- ObservedSourceData$new(case_study_config)
```

Set the observed target data (in the paediatrics population)
```{r}
target_data <- ObservedTargetData$new(treatment_effect_estimate = case_study_config$target$treatment_effect, treatment_effect_standard_error = case_study_config$target$standard_error, target_sample_size_per_arm = as.integer(case_study_config$target$total / 2), summary_measure_likelihood = case_study_config$summary_measure_likelihood)

print(target_data$sample)

n_paediatrics <- as.integer(case_study_config$target$total / 2)
```
$$
\text { Robust mixture prior }  =w \times \operatorname{Normal}\left(m_{\text {adult }}, v_{\text {adult }}\right)+(1-w) \times \operatorname{Normal}\left(m_{\text {weak }}, v_{\text {weak }}\right) 
$$ 

$$ \text { Robust mixture prior }=w \times \operatorname{Normal}(-0.694,0.017)+(1-w) \times \operatorname{Normal}(0,12.4)$$


Note that, at the moment, the code does not allow specifying different sample sizes for the arms of the target study.
 
## Inference

For each parameter, instantiate a model, perform Bayesian inference and compute the posterior mean and credible interval.

The model instantiation relies on the so-called factory design pattern for increased ease of use and flexibility.
 

```{r, message=FALSE, error=FALSE}
n_samples_posterior <- 10000000

point_est <- numeric(Nw)
credible_intervals <- matrix(NA, 2, Nw)
confidence_level <- 0.95
lower_quantile <- (1 - confidence_level) / 2
upper_quantile <- 1 - lower_quantile

posterior_w <- numeric(Nw)

for (i in seq_along(w_range)) {
  method_parameters$prior_weight <- w_range[i]

  model <- Model$new()
  model <- model$create(
    case_study_config = case_study_config,
    method = method,
    method_parameters = method_parameters,
    source_data = source_data
  )

  model$vague_prior_variance <- n_paediatrics * target_data$sample$treatment_effect_standard_error^2 # Following Best et al, 2021.

  # Perform Bayesian inference based on observed target data.
  model$inference(target_data = target_data)

  # Convert the estimates back to the natural scale
  point_est[i] <- exp(model$posterior_mean())

  # To estimate the credible intervals, we sample from the posterior distribution
  posterior_samples <- exp(model$sample_posterior(n_samples_posterior))
  credible_intervals[, i] <- quantile(posterior_samples, probs = c(lower_quantile, upper_quantile))
  posterior_w[i] <- model$wpost
}
```

### Plot results

We plot the results as in the original paper by [Best et al (2021)](https://onlinelibrary.wiley.com/doi/10.1002/pst.2093).


```{r, fig.width=5, fig.height=3}
ggplot2::ggplot() +
  geom_errorbar(
    ggplot2::aes(
      x = w_range,
      y = point_est,
      ymin = credible_intervals[1, ],
      ymax = credible_intervals[2, ]
    ),
    color = "black", linewidth = 1, width = 0
  ) +
  geom_point(ggplot2::aes(x = w_range, y = point_est), color = "black", size = 3) +
  geom_hline(yintercept = 1, linetype = "dashed", linewidth = 1) +
  ggplot2::labs(
    x = "Prior weight on adult data",
    y = "Rates ratio (95% CI, log scale)",
    title = "Posterior Mean with the Gaussian RMP"
  ) +
  scale_y_log10(limits = c(0.0625, 4), breaks = c(0.0625, 0.125, 0.25, 0.5, 1, 2, 4)) +
  theme_bw() +
  theme(
    text = element_text(size = 14),
    axis.text.y = element_text(size = 7),
    axis.ticks.y = element_blank(),
    strip.text.x = element_text(size = 10),
    legend.key = element_blank(),
    strip.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

### Table summarizing the results

[Best et al (2021)](https://onlinelibrary.wiley.com/doi/10.1002/pst.2093) presents a table summarizing the results of inference for various prior weights.

```{r}
data <- data.frame(
  "Prior weight on adult data" = seq(0, 1, by = 0.1),
  "Rate Ratio (95% CrI) (Mepo/Placebo)" = c(
    "0.68 (0.18, 2.64)",
    "0.55 (0.20, 2.33)",
    "0.53 (0.23, 2.07)",
    "0.52 (0.25, 1.85)",
    "0.51 (0.28, 1.64)",
    "0.51 (0.32, 1.44)",
    "0.51 (0.36, 1.23)",
    "0.51 (0.37, 0.99)",
    "0.51 (0.38, 0.74)",
    "0.51 (0.39, 0.67)",
    "0.50 (0.39, 0.65)"
  ),
  check.names = FALSE
)

# Create and style the table
kable(data, "html") %>%
  kableExtra::kable_styling("striped", full_width = F)
```


```{r}
Nw <- 11
w_range <- seq(0, 1, length.out = Nw)
point_est <- numeric(Nw)
credible_intervals <- matrix(NA, 2, Nw)
confidence_level <- 0.95
lower_quantile <- (1 - confidence_level) / 2
upper_quantile <- 1 - lower_quantile

posterior_w <- numeric(Nw)

for (i in seq_along(w_range)) {
  method_parameters$prior_weight <- w_range[i]

  model <- Model$new()
  model <- model$create(
    case_study_config = case_study_config,
    method = method,
    method_parameters = method_parameters,
    source_data = source_data
  )
  model$vague_prior_variance <- n_paediatrics * target_data$sample$treatment_effect_standard_error^2 # Following Best et al, 2021.

  # Perform Bayesian inference based on observed target data.
  model$inference(target_data = target_data)

  # Convert the estimates back to the natural scale
  point_est[i] <- exp(model$posterior_mean())

  # To estimate the credible intervals, we sample from the posterior distribution
  posterior_samples <- exp(model$sample_posterior(n_samples_posterior))
  credible_intervals[, i] <- quantile(posterior_samples, probs = c(lower_quantile, upper_quantile))
  posterior_w[i] <- model$wpost
}


point_est <- round(point_est, 2)
credible_intervals <- round(credible_intervals, 2)

data <- data.frame(
  `Prior weight on adult data` = w_range,
  `Rate Ratio (95% CrI) (Mepo/Placebo)` = paste(
    point_est,
    paste0("(", round(credible_intervals[1, ], 2), ", ", round(credible_intervals[2, ], 2), ")")
  ),
  check.names = FALSE
)
# Create and style the table
kable(data, "html", escape = FALSE) %>%
  kableExtra::kable_styling("striped", full_width = F)
```

  
### Posterior weight vs prior weight

[Best et al (2021)](https://onlinelibrary.wiley.com/doi/10.1002/pst.2093) contains a figure representing the posterior weight on adult data as a function of the prior weight on adult data. 
```{r} 
data <- data.frame(
  w = w_range,
  posterior_w = posterior_w
)

scatter_plot <- ggplot2::ggplot(data, ggplot2::aes(x = w_range, y = posterior_w)) +
  geom_point() +
  ggtitle("Prior versus posterior weight on adult component of the robust mixture prior") +
  xlab("Prior weight on adult data") +
  ylab("Posterior weight on adult data") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

print(scatter_plot)
```

## Comparison with RBesT

```{r}
Nw <- 21
w_range <- seq(0, 1, length.out = Nw)
point_est <- numeric(Nw)
credible_intervals <- matrix(NA, 2, Nw)
confidence_level <- 0.95
lower_quantile <- (1 - confidence_level) / 2
upper_quantile <- 1 - lower_quantile

posterior_w <- numeric(Nw)

for (i in seq_along(w_range)) {
  prior.sri <- RBesT::mixnorm(c(w_range[i], source_data$treatment_effect_estimate, sqrt(0.017)), c(1 - w_range[i], 0, sqrt(12.4)))
  post.sri <- RBesT::postmix(prior.sri, m = target_data$sample$treatment_effect_estimate, se = target_data$sample$treatment_effect_standard_error)
  res.sri <- summary(post.sri)

  # Convert the estimates back to the natural scale
  point_est[i] <- exp(res.sri["mean"])

  credible_intervals[, i] <- c(exp(res.sri[3]), exp(res.sri[5]))
}
```

```{r, fig.width=5, fig.height=3}
ggplot2::ggplot() +
  geom_errorbar(
    ggplot2::aes(
      x = w_range,
      y = point_est,
      ymin = credible_intervals[1, ],
      ymax = credible_intervals[2, ]
    ),
    color = "black", linewidth = 1, width = 0
  ) +
  geom_point(ggplot2::aes(x = w_range, y = point_est), color = "black", size = 3) +
  geom_hline(yintercept = 1, linetype = "dashed", linewidth = 1) +
  ggplot2::labs(
    x = "Prior weight on adult data",
    y = "Rates ratio (95% CI, log scale)",
    title = "Posterior Mean with the Gaussian RMP"
  ) +
  scale_y_log10(limits = c(0.0625, 4), breaks = c(0.0625, 0.125, 0.25, 0.5, 1, 2, 4)) +
  theme_bw() +
  theme(
    text = element_text(size = 14),
    axis.text.y = element_text(size = 7),
    axis.ticks.y = element_blank(),
    strip.text.x = element_text(size = 10),
    legend.key = element_blank(),
    strip.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```
```{r}
Nw <- 11
w_range <- seq(0, 1, length.out = Nw)
point_est <- numeric(Nw)
credible_intervals <- matrix(NA, 2, Nw)
confidence_level <- 0.95
lower_quantile <- (1 - confidence_level) / 2
upper_quantile <- 1 - lower_quantile

posterior_w <- numeric(Nw)

for (i in seq_along(w_range)) {
  prior.sri <- RBesT::mixnorm(c(w_range[i], source_data$treatment_effect_estimate, sqrt(0.017)), c(1 - w_range[i], 0, sqrt(12.4)))
  post.sri <- RBesT::postmix(prior.sri, m = target_data$sample$treatment_effect_estimate, se = target_data$sample$treatment_effect_standard_error)
  res.sri <- summary(post.sri)


  model$vague_prior_variance <- n_paediatrics * target_data$sample$treatment_effect_standard_error^2 # Following Best et al, 2021.

  # Perform Bayesian inference based on observed target data.
  model$inference(target_data = target_data)

  # Convert the estimates back to the natural scale
  point_est[i] <- exp(res.sri["mean"])

  # To estimate the credible intervals, we sample from the posterior distribution
  posterior_samples <- exp(RBesT::rmix(post.sri, n = n_samples_posterior))
  credible_intervals[, i] <- quantile(posterior_samples, probs = c(lower_quantile, upper_quantile))
}

point_est <- round(point_est, 2)
credible_intervals <- round(credible_intervals, 2)

data <- data.frame(
  `Prior weight on adult data` = w_range,
  `Rate Ratio (95% CrI) (Mepo/Placebo)` = paste(
    point_est,
    paste0("(", round(credible_intervals[1, ], 2), ", ", round(credible_intervals[2, ], 2), ")")
  ),
  check.names = FALSE
)
# Create and style the table
kable(data, "html", escape = FALSE) %>%
  kableExtra::kable_styling("striped", full_width = F)
```

## Other treatment effect values in the target study

In Figure 5 of [Best et al (2021)](https://onlinelibrary.wiley.com/doi/10.1002/pst.2093), several other values are considered for the RR in adolescents (2, 1.25, 0.9, 0.3):
```{r frequentist_ocs, message=FALSE, error=FALSE}
target_data$sample$treatment_effect_estimate <- log(2)


Nw <- 20
point_est <- numeric(Nw)
w_range <- seq(0, 1, length.out = Nw)
credible_intervals <- matrix(NA, 2, Nw)

posterior_w <- numeric(Nw)

for (i in seq_along(w_range)) {
  method_parameters$prior_weight <- w_range[i]

  model <- Model$new()
  model <- model$create(
    case_study_config = case_study_config,
    method = method,
    method_parameters = method_parameters,
    source_data = source_data
  )

  model$vague_prior_variance <- n_paediatrics * target_data$sample$treatment_effect_standard_error^2 # Following Best et al, 2021.

  # Perform Bayesian inference based on observed target data.
  model$inference(target_data = target_data)

  # Convert the estimates back to the natural scale
  point_est[i] <- exp(model$posterior_mean())

  # To estimate the credible intervals, we sample from the posterior distribution
  posterior_samples <- exp(model$sample_posterior(n_samples_posterior))
  credible_intervals[, i] <- quantile(posterior_samples, probs = c(lower_quantile, upper_quantile))
  posterior_w[i] <- model$wpost
}
```

## Plot results

We plot the results as in the original paper by [Best et al (2021)](https://onlinelibrary.wiley.com/doi/10.1002/pst.2093)

```{r, fig.width=5, fig.height=3}
ggplot2::ggplot() +
  geom_errorbar(
    ggplot2::aes(
      x = w_range,
      y = point_est,
      ymin = credible_intervals[1, ],
      ymax = credible_intervals[2, ]
    ),
    color = "black", linewidth = 1, width = 0
  ) +
  geom_point(ggplot2::aes(x = w_range, y = point_est), color = "black", size = 3) +
  geom_hline(yintercept = 1, linetype = "dashed", linewidth = 1) +
  ggplot2::labs(
    x = "Prior weight on adult data",
    y = "Rates ratio (95% CI, log scale)",
    title = "Posterior Mean with the Gaussian RMP"
  ) +
  scale_y_log10(limits = c(0.0625, 8), breaks = c(0.0625, 0.125, 0.25, 0.5, 1, 2, 4, 8)) +
  theme_bw() +
  theme(
    text = element_text(size = 14),
    axis.text.y = element_text(size = 7),
    axis.ticks.y = element_blank(),
    strip.text.x = element_text(size = 10),
    legend.key = element_blank(),
    strip.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

Comparison with RBesT: 
```{r}
Nw <- 20
w_range <- seq(0, 1, length.out = Nw)
point_est <- numeric(Nw)
credible_intervals <- matrix(NA, 2, Nw)
confidence_level <- 0.95
lower_quantile <- (1 - confidence_level) / 2
upper_quantile <- 1 - lower_quantile

posterior_w <- numeric(Nw)

for (i in seq_along(w_range)) {
  prior.sri <- RBesT::mixnorm(c(w_range[i], source_data$treatment_effect_estimate, sqrt(0.017)), c(1 - w_range[i], 0, sqrt(12.4)))
  post.sri <- RBesT::postmix(prior.sri, m = target_data$sample$treatment_effect_estimate, se = target_data$sample$treatment_effect_standard_error)
  res.sri <- summary(post.sri)

  # Convert the estimates back to the natural scale
  point_est[i] <- exp(res.sri["mean"])

  credible_intervals[, i] <- c(exp(res.sri[3]), exp(res.sri[5]))
}
```

```{r plot_results, fig.width=5, fig.height=3}
ggplot2::ggplot() +
  geom_errorbar(
    ggplot2::aes(
      x = w_range,
      y = point_est,
      ymin = credible_intervals[1, ],
      ymax = credible_intervals[2, ]
    ),
    color = "black", linewidth = 1, width = 0
  ) +
  geom_point(ggplot2::aes(x = w_range, y = point_est), color = "black", size = 3) +
  geom_hline(yintercept = 1, linetype = "dashed", linewidth = 1) +
  ggplot2::labs(
    x = "Prior weight on adult data",
    y = "Rates ratio (95% CI, log scale)",
    title = "Posterior Mean with the Gaussian RMP"
  ) +
  scale_y_log10(limits = c(0.0625, 8), breaks = c(0.0625, 0.125, 0.25, 0.5, 1, 2, 4, 8)) +
  theme_bw() +
  theme(
    text = element_text(size = 14),
    axis.text.y = element_text(size = 7),
    axis.ticks.y = element_blank(),
    strip.text.x = element_text(size = 10),
    legend.key = element_blank(),
    strip.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

Posterior and prior pdf : 
```{r}
method_parameters$prior_weight <- 0.5

model <- Model$new()
model <- model$create(
  case_study_config = case_study_config,
  method = method,
  method_parameters = method_parameters,
  source_data = source_data
)

# Perform Bayesian inference based on observed target data.
model$inference(target_data = target_data)
model$plot_pdfs(xmin = -1, xmax = 1, resolution = 100)
```
 
